---
title: "Stratigraphic Plots in ggplot2"
author: "Richard J. Telford"
date: "May 18, 2018"
output: 
  html_document:
    self_contained: false
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
library("neotoma")
library("tidyverse", quietly = TRUE)
```

`rioja::strat.plot` is a great tool for plotting stratigraphic plots in R, but sometimes it is not obvious how to do something I want, perhaps a summary panel showing the percent trees/shrubs/herbs. Of course, I could extend `strat.plot`, but I do all nearly all my figures with `ggplot` now and wouldn't know where to start.

In this post, I'm going to show how to make a stratigraphic plot with `ggplot`.

First I'm going to download the pollen data for Tsuolbmajavri (Seppä and Hicks, 2006) from Neotoma.

```{r download}
library("neotoma")
library("tidyverse")

#get dataset list
siteID <- 15733
pollen_sites <- get_dataset(datasettype = "pollen", x = siteID)

#download_data
pollen_data <- get_download(pollen_sites)
pollen_data <- pollen_data[[1]]

#species groups to plot
wanted <- c("TRSH", "UPHE", "VACR", "SUCC", "PALM", "MANG")

```

Now I need to extract and combine the pollen count, the chronology and the taxon information from the downloaded neotoma data. I `gather` the data to make it into a long thin (tidy) table, filter the taxa that are in the ecological groups I want to plot and calculate percentages.  

```{r process_data}
thin_pollen <- bind_cols(
  counts(pollen_data),
  ages(pollen_data) %>% select(depth, age, age.type)
) %>% 
  gather(key = taxon, value = count, -depth, -age, -age.type) %>% 
  left_join(pollen_data$taxon.list, c("taxon" = "taxon.name")) %>% 
  filter(ecological.group %in% wanted) %>% 
  mutate(ecological.group = factor(ecological.group, levels = wanted)) %>% 
  group_by(age) %>% 
  mutate(percent = count / sum(count) * 100)

```

Now I can plot the data after a little further processing: I'm selecting taxa that occur at least three times and with a maximum abundance of at least 3%, and making taxon into a factor so taxa from the same ecological group will plot together.

I use `geom_area` twice, once to show the 10x the percent, once to show the percent. I then need to use `ylim` in `coord_flip` to truncate the axes at the maximum pollen abundance. The plot uses facets to show the taxa separately. 

```{r plot}
plot_data <- thin_pollen %>% 
  mutate(taxon = factor(taxon, levels = unique(taxon[order(ecological.group)]))) %>% 
  group_by(taxon) %>% 
  filter(max(percent) >= 3, sum(percent > 0) >= 3) %>% 
  ungroup() %>% 
  mutate(shaddow = percent * 10, shaddow = if_else(shaddow > max(percent), max(percent), shaddow))
  
g <- ggplot(plot_data, aes(x = age, y = percent, fill = ecological.group)) +
  geom_area(aes(y = shaddow), alpha = 0.3, show.legend = FALSE) +
  geom_area(show.legend = FALSE) +
  coord_flip() +
  scale_x_reverse(expand = c(0.02, 0)) +
  scale_y_continuous(expand = c(0, 0), breaks = scales::pretty_breaks(n = 2)) +
  scale_fill_brewer(type = "qual", palette = "Set1", drop = FALSE) +
  facet_grid(~ taxon, scales = "free", space = "free") +
  labs(x = expression(Age~''^{14}*C~BP), y = "Percent", fill = "Group") + 
  theme(strip.text.x = element_text(angle = 90, size = 7, hjust = 0, margin = unit(rep(1, 4), unit = "pt")), 
        panel.spacing = unit(x = 1, units = "pt"),
        axis.text.x = element_text(size = 7)
        )
g
```


I don't think that is too bad for a first attempt, but I think I'll stick to `rioja::strat.plot` most of the time. 




```{r summary}
summary_data <- thin_pollen %>% 
  group_by(age, ecological.group) %>% 
  summarise(sum = sum(percent)) %>% 
  mutate(ecological.group = factor(ecological.group), 
         ecological.group = factor(ecological.group, levels = wanted), 
         summary = "Summary") 

summary_plot <- summary_data %>% 
  ggplot(aes(x = age, y = sum, fill = ecological.group, group = ecological.group)) +
  geom_area(position = position_stack(reverse = TRUE), show.legend = FALSE) +
  scale_x_reverse(expand = c(0.02, 0)) +
  scale_y_continuous(expand = c(0.01, 0)) +
  scale_fill_brewer(type = "qual", palette = "Set1", drop = FALSE) + 
  coord_flip() +
  facet_wrap(~summary)

no_x <- theme(axis.text.y = element_blank(),
              axis.title.y = element_blank(),
              axis.ticks.y = element_blank(),
              plot.margin = unit(x = c(5.5, 5.5, 5.5, 0), units = "pt"), 
              strip.text.x = element_text(angle = 90, size = 7, hjust = 0, vjust = 0, margin = unit(rep(1, 4), unit = "pt")),
              axis.text.x = element_text(size = 7),
              axis.title.x = element_blank())

cowplot::plot_grid(g + theme(plot.margin = unit(x = c(5.5, 1, 5.5, 5.5), units = "pt")), 
                   summary_plot + no_x, align = "h", rel_widths = c(85, 15), axis = c("t", "b"))
```

```{r sed_units, eval = FALSE}
FaciesTypes <- get_table("FaciesTypes")
AnalysisUnits <- get_table("AnalysisUnits")
CollectionUnits <- get_table("CollectionUnits")
```

```{r, eval = FALSE}
CollectionUnits %>%
  filter(SiteID == siteID) %>% 
  left_join(AnalysisUnits)

```

```{r}
data_frame(unit = c("a", "b", "a", "c"), upper = c(0, 1, 1.2, 2), lower =c(1, 1.2, 2, 2.5)) %>% 
  ggplot(aes(xmin = -Inf, xmax = Inf, ymin = lower, ymax = upper, fill = unit)) +
  geom_rect() +
  expand_limits(x = c(0, 1)) +
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

```

